An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2013-11-19 paper.tex
Re-named Figures to FigureNN_name
Dan's edits (11/8/2012 email: Paper (First Draft))

\newcommand{\vtheta}{\vec{\theta}}

@@

\author{W M Farr$^1$ and I Mandel$^2$}

\address{$^1$ Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA) \\ Department of Physics and Astronomy, Northwestern University}

\address{$^2$ MIT Kavli Institute; and University of Birmingham, UK}

\eads{\mailto{w-farr@northwestern.edu}, \mailto{ilyamandel@chgk.info}}

\begin{abstract}
  Selection among alternative theoretical models given an observed
  data set is an important challenge in many areas of physics and
  astronomy.  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an
  extremely powerful technique for performing Bayesian model
  selection, but it suffers from a fundamental difficulty: it requires
  jumps between model parameter spaces, but cannot retain a memory of
  the favored locations in more than one parameter space at a time.
  Thus, a naive jump between parameter spaces is unlikely to be
  accepted in the MCMC algorithm and convergence is correspondingly
  slow.  Here we demonstrate an interpolation technique that uses
  samples from single-model MCMCs to propose inter-model jumps from an
  approximation to the single-model posterior of the target parameter
  space.  The interpolation technique, based on a kD-tree data
  structure, is adaptive and efficient in arbitrary dimensions.  We
  show that our technique leads to dramatically improved convergence
  over naive jumps in an RJMCMC, and compare it to other proposals in
  the literature to improve the convergence of RJMCMCs.  We also
  discuss the use of the same interpolation technique in two other
  contexts: as a convergence test for a single-model MCMC and as a way
  to construct efficient ``global'' proposal distributions for
  single-model MCMCs without prior knowledge of the structure of the
  posterior distribution.
\end{abstract}

@@

Here we introduce an alternative technique for improving the
acceptance ratio of inter-model jumps in an RJMCMC, leading to
dramatically improved convergence of RJMCMC sample chains.  The
technique uses a kD-tree data structure to construct an approximation
to each model's posterior parameter distribution.  We draw jump
proposals into the model from this approximation to its posterior.
Because jumps are proposed preferentially to locations favored by the
single-model posterior, the RJMCMC compares ``good'' locations in
parameter space across all the models, and convergence is generally
rapid.  We have successfully applied this RJMCMC technique to a 10-way
model selection among alternative mass distribution models for
black-hole X-ray binaries \cite{Farr2010}.

The method of \cite{Littenberg2009} for producing inter-model jumps in
an RJMCMC relies on a box decomposition of parameter space, using
fixed-sized boxes.  The method cannot adapt to the local structure of
the posterior, and becomes asymptotically inefficient for
high-dimensional parameter spaces or highly peaked posteriors.
Meanwhile, the approximation to the posterior distribution produced by
the kD-tree is a constant-in-box interpolation of the posterior,
similar in spirit to the phase-space density interpolants produced
from N-body positions and momenta in \cite{Ascasibar2005}.  The
kD-tree interpolation is effective in parameter spaces of arbitrary
dimensionality, and is quite space-efficient, requiring $\order{N}$
storage space and $\order{\log N}$ time to produce each proposed jump,
where $N$ is the number of single-model MCMC samples used to construct
the interpolation.

The structure of this paper is as follows.  In Section
\ref{sec:reversible-jump} we introduce in more detail the concept of a
Reversible Jump MCMC, and describe the fundamental difficulty with a
naive jump proposal in an RJMCMC.  In Section \ref{sec:kDTree} we
introduce the kD-tree data structure used to decompose the parameter space
into boxes for interpolation.  In Section \ref{sec:efficiency} we
demonstrate the efficiency gains that are achieved from use of the
interpolated jump proposal.  In Section \ref{sec:examples} we give
examples of some other uses of the interpolated jump proposal, and suggest its utility in the context of a
single-model MCMC.  Finally, in Section \ref{sec:conclusion}
we offer a summary and some concluding remarks on the method.

@@

The jump proposal is the most important choice in the MCMC, as it
determines the sampling efficiency of the algorithm, i.e., the length
of the chain before it converges to the posterior PDF.  Creating an
efficient jump proposal distribution requires an understanding of the
structure of the parameter space which may not be available until the
PDFs are found, creating a Catch-22; one possibility for resolving
this infinite loop is described in Section \ref{sec:examples}.

It should be noted that although an MCMC whose jump acceptance
criterium obeys detailed balance (as the Metropolis-Hastings algorithm
does) must eventually converge to the desired distribution, there is
no way to guarantee convergence in a fixed number of steps or to test
whether a chain has converged in a foolproof manner.  For example,
MCMC chains can get stuck on local maxima, producing an apparently
well-converged sampling of the PDF in the vicinity of the maximum; or,
if the chain visits a sequence of local maxima, moving rarely between
maxima, the autocorrelation length of the chain may represent a
substantial fraction of the total number of samples, resulting in an
effective sample size that is too small to accurately represent the
relative sizes of the modes in the PDF (however, see Section
\ref{sec:examples} for one intriguing suggestion for remedying this
issue).

@@

Here we propose to use a decomposition of the parameter space into
neighborhoods based on a data structure called a kD-tree (see, e.g.\
\cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
efficient in high-dimensional spaces than the Voronoi decomposition.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{kdtree}
  \end{center}
  \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
    constructed around a set of points that are normally distributed
    about the origin in two dimensions.  As the points become denser
    around the origin, the typical neighborhood gets smaller.  The
    interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
    where $N$ is the total number of points (which is also the number
    of boxes).}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{acceptRate}
  \end{center}
  \caption{\label{fig:acceptRate} The inter-model acceptance rate
    versus the number of points per box when the kD-tree neighborhood
    search is truncated.  As the number of points per box increases,
    and the interpolation becomes less accurate, the acceptance rate
    falls, asymptoting to the rate for naive draws from the uniform
    prior (about $5\%$ for this data set).}
\end{figure}

@@

\section{Examples and other uses} 
\label{sec:examples}

RJMCMC with efficient PDF interpolation via kD trees can be used in a large variety of problems in physics and astronomy, whenever Bayesian analysis is used to perform model selection.  Below, we describe several scenarios in which we have successfully applied this technique.  Moreover, PDF interpolation with a kD tree can be extremely useful in other contexts, beyond a reversible-jump MCMC.  We suggest two examples below: generating efficient jump proposal distributions in a single-model MCMC and convergence tests.

We have successfully employed the RJMCMC technique described above
when evaluating several alternative models for the distribution of the
masses of black holes in X-ray binaries \cite{Farr2010}.  We
performed a Bayesian analysis of the mass distribution of stellar-mass
black holes using the observed masses of 15 low-mass X-ray binary
systems undergoing Roche lobe overflow and five high-mass, wind-fed
X-ray binary systems.  We considered ten different mass distribution
models: Gaussian, double Gaussian, power law, exponential decay, log
normal, and 1-, 2-, 3-, 4-, and 5-bin histograms.  Each model was
described by between two and six parameters.  Model selection using
RJMCMC with kD trees allowed us to determine that the mass
distribution of the low-mass systems is best fit by a power-law, while
the distribution of the combined sample is best fit by the exponential
model.  Based on the model selection, we were able to determine which
models would provide the best information about astrophysically
relevant parameters, such as the minimum black hole mass.  We were
also able to conclude that the low-mass subsample is not consistent
with being drawn from the distribution of the combined population
\cite{Farr2010}.

Gravitational-wave astronomy provides the setting for another ongoing
study.  Interesting triggers from the LIGO \cite{InitLIGO} and Virgo
\cite{Virgo} gravitational-wave detector pipeline that searches for
compact binary coalescences are followed up with Bayesian parameter
estimation tools (see, e.g.,
\cite{Rover:2006ni,vanderSluys:2008b,VeitchVecchio:2009}).  In
general, several possible models for the data are considered,
including a pure noise model, noise superimposed on a non-spinning
gravitational-wave signal, or noise superimposed on a gravitational
wave signal that includes significant spin or spins in the binary
components.  The models can include a large number of parameters (up
to 15 for the model with two spinning components \cite{Raymond:2010}),
and RJMCMC with interpolation via kD trees is again successful at
computing the model evidences.

As discussed in Section \ref{sec:mcmc}, finding an efficient jump
proposal distribution can be a challenge even for a single-model MCMC.
This is particularly difficult when the parameter space is
multi-modal, in which case the Markov chain can get stuck on
individual high-posterior islands in the parameter space.  Traversals
of low-posterior oceans by a series of short jumps are extremely
unlikely, and random long jumps are almost always rejected, making
sampling very inefficient.  Essentially, such isolated islands in
parameter space behave as separate models, and weighting them by the
number of samples in each peak then corresponds to model selection.
Ideally, an analytical understanding of the model would allow for the
creation of jump proposal distributions that combine short jumps
designed to explore individual islands with directed long jumps that
are likely to end up on neighboring islands.  However, it is not
always possible to gain such understanding short of performing the
MCMC itself.  One solution lies in running multiple chains that
explore the parameter space, perhaps getting stuck on local maxima,
and then combining the points and interpolating with a kD tree to
create a jump proposal distribution.  Although the chains in the first
iteration will not have transitioned between the peaks, and the
samples will therefore not weight the separate peaks appropriately,
this will ensure that all of the modes (islands) are included in the
jump proposal distribution for a second stage.  This second stage can
employ an interpolated jump proposal exactly as in the model selection
discussed above.  Other modifications to this technique, which we are
currently investigating, include multiple iterative stages, with the
interpolated samples from each previous stage used as a jump proposal
distribution for the next stage, until the posterior PDF is read off
from the final stage; and the use of higher temperatures in the
early-stage chains to improve sampling.

An example of a multi-modal parameter space where chains can get stuck
on high-posterior islands comes from antipodal sky location
degeneracies for networks of three ground-based gravitational-wave
interferometers or for low-frequency gravitational wave signals in the
proposed space-borne instrument LISA \cite{LISA}.  Although PDFs
produced by MCMC chains will generally show both locations on the sky,
the chain length may not be sufficiently long in practice to include
enough jumps between the two high-likelihood antipodal locations.  In
general, the relative weight of two separated peaks is determined with
a fractional error that scales as
$1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of transitions between
the peaks.  By increasing the number of proposed transitions, the
convergence of such PDFs can be enhanced using the kD-tree
interpolated proposal.  As described above, one could test the convergence of the PDF 
by running a followup MCMC with a jump proposal
distribution from an interpolation of the first original PDF.  This
follow-up MCMC would attempt large global jump proposals rather than
small local ones, meaning that it could rapidly accumulate sufficient
points to determine whether the original run spent the correct fractional
amounts of time in the various modes.

@@

We demonstrated the efficiency of this technique on a toy model-comparison problem described in Section \ref{sec:efficiency}.  We also successfully applied this technique to the problem of selecting the best model for the observed distribution of black-hole X-ray binaries, as described in Section \ref{sec:examples} and Ref.~\cite{Farr2010}.   In addition to model comparison, the PDF interpolation described here can be useful in multi-step MCMCs to improve the sampling efficiency by selecting better jump proposal distributions or to test MCMC convergence.

We have made our implementation of the technique described in this
paper publicly available online at
\url{http://github.com/farr/mcmc-ocaml}, and welcome readers to take
advantage of this toolkit.

@@

We are grateful to Neil Cornish for interesting discussions.  IM
acknowledges support from the NSF Astronomy and Astrophysics
Postdoctoral Fellowship, award AST-0901985.  WF acknowledges support
from NSF grant AST0908930 and NASA grant NNX09AJ56G.

>>>>

\newcommand{\vtheta}{\vec{\theta}}
\newcommand{\Nbox}{N_\mathrm{box}}
\newcommand{\Vbox}{V_\mathrm{box}}

@@

\author{W M Farr$^{1,3}$, I Mandel$^{2,3}$ and D Stevens$^{1,4}$}

\address{$^1$ Center for Interdisciplinary Exploration and Research in
  Astrophysics (CIERA), Department of Physics and Astronomy,
  Northwestern University, Evanston IL USA}

\address{$^2$ MIT Kavli Institute, Cambridge MA USA}

\address{$^3$ School of Physics and Astronomy, University of
  Birmingham, Birmingham UK}

\address{$^4$ Department of Astronomy, The Ohio State
  University, Columbus OH USA}

\eads{\mailto{w.farr@bham.ac.uk}, \mailto{ilyamandel@chgk.info} \mailto{dan.stevens@u.northwestern.edu}}

\begin{abstract}
  Selection among alternative theoretical models given an observed
  data set is an important challenge in many areas of physics and
  astronomy.  Reversible-jump Markov chain Monte Carlo (RJMCMC) is an
  extremely powerful technique for performing Bayesian model
  selection, but it suffers from a fundamental difficulty: it requires
  jumps between model parameter spaces, but cannot retain a memory of
  the favored locations in more than one parameter space at a time.
  Thus, a naive jump between parameter spaces is unlikely to be
  accepted in the MCMC algorithm and convergence is correspondingly
  slow.  Here we demonstrate an interpolation technique that uses
  samples from single-model MCMCs to propose inter-model jumps from an
  approximation to the single-model posterior of the target parameter
  space.  The interpolation technique, based on a kD-tree data
  structure, is adaptive and efficient in modest dimensionality.  We
  show that our technique leads to dramatically improved convergence
  over naive jumps in an RJMCMC, and compare it to other proposals in
  the literature to improve the convergence of RJMCMCs.  We also
  demonstrate the use of the same interpolation technique as a way to
  construct efficient ``global'' proposal distributions for
  single-model MCMCs without prior knowledge of the structure of the
  posterior distribution, and discuss improvements that permit the
  method to be used in higher-dimensional spaces efficiently.
\end{abstract}

@@

Here we introduce an alternative technique for improving the
acceptance ratio of inter-model jumps in an RJMCMC, leading to
dramatically improved convergence of RJMCMC sample chains.  The
technique uses a kD-tree data structure to construct an approximation
to each model's posterior parameter distribution.  We draw jump
proposals into the model from this approximation to its posterior.
Because jumps are proposed preferentially to locations favored by the
single-model posterior, the RJMCMC compares ``good'' locations in
parameter space across all the models, and convergence is generally
rapid.  We have successfully applied this RJMCMC technique to a 10-way
model selection among alternative mass distribution models for
black-hole X-ray binaries \cite{Farr2010}.  We also provide an example
using this method as an ``asymptotically Markovian''
\cite{terBraak2008} jump proposal in the context of a single-model,
nine-dimensional MCMC in Section \ref{sec:higherdim}.

The method of \cite{Littenberg2009} for producing inter-model jumps in
an RJMCMC relies on a box decomposition of parameter space, using
fixed-sized boxes.  The method cannot adapt to the local structure of
the posterior, and becomes asymptotically inefficient for
high-dimensional parameter spaces or highly peaked posteriors.
Meanwhile, the approximation to the posterior distribution produced by
the kD-tree is a constant-in-box interpolation of the posterior,
similar in spirit to the phase-space density interpolants produced
from N-body positions and momenta in \cite{Ascasibar2005}.  The
kD-tree interpolation is effective in parameter spaces of modest
dimensionality, and is quite space-efficient, requiring $\order{N}$
storage space and $\order{\log N}$ time to produce each proposed jump,
where $N$ is the number of single-model MCMC samples used to construct
the interpolation.

The structure of this paper is as follows.  In Section
\ref{sec:reversible-jump} we introduce in more detail the concept of a
Reversible Jump MCMC, and describe the fundamental difficulty with a
naive jump proposal in an RJMCMC.  In Section \ref{sec:kDTree} we
introduce the kD-tree data structure used to decompose the parameter
space into boxes for interpolation.  In Section \ref{sec:efficiency}
we demonstrate the efficiency gains that are achieved from use of the
interpolated jump proposal.  In Section \ref{sec:higherdim} we give
examples of some other uses of the interpolated jump proposal that
suggest its utility in the context of a single-model MCMC.  Finally,
in Section \ref{sec:conclusion} we offer a summary and some concluding
remarks on the method.

@@

The jump proposal is the most important choice in the MCMC, as it
determines the sampling efficiency of the algorithm, i.e., the length
of the chain before it converges to the posterior PDF.  Creating an
efficient jump proposal distribution requires an understanding of the
structure of the parameter space which may not be available until the
PDFs are found, creating a Catch-22; one possibility for resolving
this infinite loop is described in Section \ref{sec:higherdim}.

It should be noted that although an MCMC whose jump acceptance
criterium obeys detailed balance (as the Metropolis-Hastings algorithm
does) must eventually converge to the desired distribution, there is
no way to guarantee convergence in a fixed number of steps or to test
whether a chain has converged in a foolproof manner.  For example,
MCMC chains can get stuck on local maxima, producing an apparently
well-converged sampling of the PDF in the vicinity of the maximum; or,
if the chain visits a sequence of local maxima, moving rarely between
maxima, the autocorrelation length of the chain may represent a
substantial fraction of the total number of samples, resulting in an
effective sample size that is too small to accurately represent the
relative sizes of the modes in the PDF.

@@

Here we propose to use a decomposition of the parameter space into
neighborhoods based on a data structure called a kD-tree (see,
e.g.\ \cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the boxes of Ref.~\cite{Littenberg2009}, and more
efficient in high-dimensional spaces than the Voronoi decomposition.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{Figure1_kdtree}
  \end{center}
  \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
    constructed around a set of points that are normally distributed
    about the origin in two dimensions.  As the points become denser
    around the origin, the typical neighborhood gets smaller.  The
    interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
    where $N$ is the total number of points (which is also the number
    of boxes).}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{Figure2_acceptRate}
  \end{center}
  \caption{\label{fig:acceptRate} The inter-model acceptance rate
    versus the number of points per box when the kD-tree neighborhood
    search is truncated.  As the number of points per box increases,
    and the interpolation becomes less accurate, the acceptance rate
    falls, asymptoting to the rate for naive draws from the uniform
    prior (about $5\%$ for this data set).}
\end{figure}

@@

\section{kD-Interpolated Jump Proposal in Higher-Dimensional
  Single-Model MCMCs}
\label{sec:higherdim}

In the model selection example from Section \ref{sec:efficiency}, the
models have two-dimensional parameter spaces; in \cite{Farr2010} the
highest-dimensional model had five dimensions.  As the number of
dimensions increases the interpolation becomes more difficult for two
reasons.  First, the number of sample points required for a given
number of subdivisions in each dimension grows exponentially with
dimension, so for reasonable numbers of points a higher-dimensional kD
tree will have fewer subdivisions along each dimension.  Second, as
the dimensionality increases the volume ``at the corners'' of each kD
tree box becomes more significant; this problem is even more
pronounced when the distribution of points does not align with the
coordinate axes along which the tree subdivides.  In this section, we
discuss practical solutions to this issue in the context of a
single-model MCMC, concluding with an example of a successful
application to gravitational-wave parameter estimation.

To address the arbitrary orientation of points, the modified jump
proposal does not descend to the leaf-nodes of the tree, but instead
stops descending whenever the number of points in the box falls below
a threshold, $\Nbox$.  We then find the principal axes of the points
contained in this box and draw from a modified box which is aligned
with these axes and whose center is the mean of the points contained
within the rectangular box. The new point is chosen uniformly from
within the bounds of the modified box, subject to the restriction that
it also lie within the original, coordinate-aligned box.  Figure
\ref{fig:PCC} illustrates this two-box method.

\begin{figure}
  \includegraphics[width=0.8\columnwidth]{PCC}
  \caption{\label{fig:PCC} A two-dimensional illustration of the two
    boxes involved in the modified kD proposal.  The larger box
    aligned with the axes is the normal kD-tree box containing the
    given points, while the tighter box is the box from which the
    modified kd-interpolated jump proposal draws a point. }
\end{figure}

The forward jump probability for this jump proposal is given by
%
\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N \left(V \cap \Vbox \right)} , \ee
%
where N is the total number of points in the kD-tree, $\Nbox$ is the
number of points in the box containing $\vtheta_{i+1}$, $V$ is the
volume of the original box containing $\vtheta_{i+1}$ and $\Vbox$ is
the volume of the principal-component box containing $\vtheta_{i+1}$.

The backward jump probability is computed in a similar fashion by
finding the principal-component box that would contain the current set
of parameters;.
%
\bel{modbackward} P(\vtheta_{i+1} \rightarrow \vtheta_{i}) =
\frac{\Nbox'}{N\left(V' \cap \Vbox'\right)} , \ee
%
where $\Nbox'$ is the number of points in the original box containing
$\vtheta_i$, $V'$ is the volume of the kD box containing $\vtheta_i$,
and $\Vbox'$ is the volume of the principal-axes box containing
$\vtheta_i$.  Note, however, that the current parameters may lie in a
region of the kD box that does not intersect the principal axes box;
in this case, the backward jump probability is zero, and the algorithm
rejects the proposed step.  When used as the only jump proposal for an
RJMCMC, it is important that the kD tree proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used as one of several proposals
in the context of an MCMC, this is no longer a requirement.

It can also be advantageous, particularly in high dimensional spaces,
to accumulate more points into the kD tree as the MCMC explores more
of the parameter space.  Because this introduces a dependence on the
previous states of the MCMC, the jump proposal is no longer
Markovian.  However, it remains \emph{asymptotically} Markovian, which
ensures that the equilibrium distribution of the MCMC is correct.  To
efficiently insert points into the tree as the chain accumulates them,
the algorithm from Section \ref{sec:kDTree} must be modified:
\begin{enumerate}
\item Instead of partitioning the point set at its median, we now
  partition the bounding box at its geometric center along a
  particular dimension, which cycles as we descend the tree levels.
  Note that this allows for empty boxes if all the points cluster to
  one side of the domain.
\item When inserting a point, we descend the tree to a leaf node,
  which now can be either empty or contain one point.  If it is empty,
  we insert the point at this node.  If it contains one point, we
  subdivide the appropriate dimension, and place both points into the
  two sub-boxes.  If both land in the same sub-box, we continue
  subdividing, until each box in the sub-sub-sub... tree contains one
  or zero points.
\end{enumerate}
Insertion of a point is $\order{\log N}$, since it involves a constant
amount of work per tree level, and there are $\log N$ tree levels.
Neighborhood lookup proceeds as before, and is therefore also
$\order{\log N}$.

We have implemented this modified kD tree proposal as one of many jump
proposals in the \texttt{LALInferenceMCMC} sampler
\cite{Raymond2009,Raymond:2010,Raymond2012}.
\texttt{LALInferenceMCMC} is a MCMC code, based on the LIGO algorithms
library, designed to sample the posterior on parameters of
compact-object binaries (masses, orbital orientation, distance, sky
position, etc) implied by their gravitational wave signals in
Earth-based gravitational wave detectors like LIGO.  The simplest such
signal has a nine-dimensional parameter space, and the posterior often
includes near-degeneracies and multiple modes that make convergence of
the MCMC chain very slow with traditional proposals.  Figure
\ref{fig:accratio} shows the acceptance ratio for the kD tree jump
proposal in such a simulation.  In spite of the very low acceptance
rate of the proposal, applying the kD tree proposal to one out of
twenty proposed jumps improved the convergence time of the simulation
by a factor of two compared to the standard suite of proposals because
it is particularly efficient at producing ``mode-hopping'' jumps,
which are difficult to produce with the other proposals in
\texttt{LALInferenceMCMC}.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{linearlog3}
  \end{center}
  \caption{\label{fig:accratio} The cumulative parameter acceptance
    ratio for the modified kD-interpolated jump proposal used in the
    \texttt{LALInferenceMCMC} code as a function of the number of
    steps (in hundreds). In spite of the small acceptance ratio, the
    kD tree proposal, applied to one in twenty proposed jumps,
    improved the convergence time of the sampler by a factor of two
    compared to using only the standard suite of proposals.}
\end{figure}

@@

We demonstrated the efficiency of this technique on a toy
model-comparison problem described in Section \ref{sec:efficiency}.
We also successfully applied this technique to the problem of
selecting the best model for the observed distribution of black-hole
X-ray binaries, as described in Ref.~\cite{Farr2010}.  In addition to
model comparison, the PDF interpolation described here can be useful
in multi-step MCMCs to improve the sampling efficiency by selecting
better jump proposal distributions or to test MCMC convergence.

We have made our implementation of the technique described in this
paper publicly available online at
\url{http://github.com/farr/mcmc-ocaml}, and also in the
\texttt{LALInferenceMCMC} sampler, at
\url{https://www.lsc-group.phys.uwm.edu/daswg/projects/lalsuite.html}.
We welcome readers to take advantage of this toolkit.

@@

We are grateful to Neil Cornish for interesting discussions.  IM
acknowledges support from the NSF Astronomy and Astrophysics
Postdoctoral Fellowship, award AST-0901985.  WF acknowledges support
from NSF grant AST0908930 and NASA grant NNX09AJ56G. DS acknowledges
support from a Northwestern University Summer Undergraduate Research Grant.
